Synthesizing Quotient Lenses

### 2018-03-07 papers/QuotientLenses/main.tex
Polishing in 3.3 -- still holding no locks

The net effect of these design decisions means that each QRE $c$ enables
us to simultaneously express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Observe that since $\canonizer(c)$ computes $\canonizer(c)(w)$ as the unique
$k$ in $\mathcal{L}(K(c))$ such that $w$ is equivalent to $k$ mod $\eqrel{c}$,
then the equivalence classes of $\eqrel{c}$ are the same as the
fibres\bcp{this is not a very familiar term for lots of people; if we're
  going to use it a lot, we should define it more carefully} of
$\canonizer(c)$.

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity}, which arises when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}~\cite{Sippu1988}, and is necessary, firstly because it
ensures that the canonizing function of a QRE is well-defined.

For example consider the QRE

@@

We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.
%
To this end, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.

We say that a regular expression $R$ is \textit{strongly unambiguous} if and
only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1, S_2$
strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

@@

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence rule for
concatenation is as follows:
\begin{prooftree}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\TrinaryInfC{$\wf{c \cdot c'}$}
\end{prooftree}

@@

\noindent The second hypothesis for the \textit{permutation} rule says that for
any two different permutations $\sigma$ and $\theta$, the languages
$W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot W(c_{\sigma(n)})$ and
$W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot W(c_{\theta(n)})$ must be
disjoint. This is because any string that matches the \textit{permutation}
combinator matches the language $W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\theta(n)})$ for some permutation $\theta$, so we require that these
regular expressions be disjoint.

>>>>

The net effect of these design decisions means that each QRE $c$ enables
us to simultaneously express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Observe that, since $\canonizer(c)$ computes $\canonizer(c)(w)$ as the unique
$k$ in $\mathcal{L}(K(c))$ such that $w$ is equivalent to $k$ mod $\eqrel{c}$,
it follows that the equivalence classes of $\eqrel{c}$ are the same as the
fibres\bcp{this is not a very familiar term for lots of people; if we're
  going to use it a lot, we should define it more carefully} of
$\canonizer(c)$.

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity}, which arises when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions\bcp{The following list is hard to
  parse.  A general rule is: Always separate adjacent math expressions by
  something more than just punctuation.} $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way.

This  condition is called
{\em strong unambiguity}~\cite{Sippu1988}, and it is necessary,
firstly,\bcp{is there a ``secondly'' coming?  I don't see it} because it
ensures that the canonizing function of a QRE is well-defined.
%
For example consider the QRE\bcp{Displayed math should be punctuated either
  (1) exactly as if it were inline (i.e., this would need a period at the
  end) or (2) with a sentence terminator (a colon or period) beforehand.  In
this case, (1) would be confusing (since . is also part of QRE syntax), so
(2) is preferable.}

@@

We also require that the regular combinators applied to
the kernels of QREs be unambiguous, since the underlying\bcp{?} lenses will
end up operating on these kernels, and these lenses impose the same
restrictions for similar reasons.
%
To this end\bcp{which end?}, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$, if, for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if, for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.

We say that a regular expression $R$ is \textit{strongly unambiguous} when
(1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1, S_2$
strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

@@

Figure~\ref{fig:qrerules} gives the inference rules for deriving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence\bcp{check spelling} rule for
concatenation is as follows:\bcp{Rather than repeating rules inline, give
  them names in the figure.}
\begin{prooftree}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\TrinaryInfC{$\wf{c \cdot c'}$}
\end{prooftree}

@@

\noindent The second hypothesis says that, for
any two different permutations $\sigma$ and $\theta$, the languages
$W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot W(c_{\sigma(n)})$ and
$W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot W(c_{\theta(n)})$ must be
disjoint. This is because any string that matches the \textit{permutation}
combinator matches the language $W(c_{\theta(1)}) \cdot W(c) \cdot \ldots \cdot
W(c_{\theta(n)})$ for some permutation $\theta$, so we require that these
regular expressions be disjoint.
